---
title: "Taxi Outline"
author: "Christoph, Chris, Carol"
date: "4/19/2017"
output: html_document
fontsize: 8pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**2015 Yellow Taxi Trip Data - NYC Open Data**
https://data.cityofnewyork.us/Transportation/2015-Yellow-Taxi-Trip-Data/ba8s-jw6u
Data set contains a random 10% of taxi rides on May 6th paid with credit card. 
**Research Questions**
We focus exploring 3 features: time (`dropoff_datetime`, `pickup_datetime`), 
location (`dropoff_latitude`, `dropoff_longitude`, `pickup_latitude`, 
`pickup_longitude`), tips (`tip_amount`). 


```{r, echo=FALSE}
taxi <- read.csv("taxi_data.csv", as.is = TRUE)
```


```{r, echo = FALSE, meassage = FALSE, warning = FALSE, include = FALSE}
library(dplyr)
library(ggplot2)
library(leaps)
library(lubridate)
library(MASS)
library(ggmap)
library(mapproj)
library(gridExtra)
library(grid)
```


```{r, echo=FALSE}
# Create `tip_pct`
taxi <- taxi %>% mutate(tip_pct = (tip_amount / fare_amount) * 100)

# Remove observation where tip_pct > 100
taxi <- taxi[-c(which(taxi$tip_pct == "NaN"), which(taxi$tip_pct == "Inf")),]
taxi <- taxi[taxi$tip_pct <= 50,]
```

```{r, echo=FALSE}
# Create `pickup_hour` and `dropoff_hour`
taxi$pickup_hour <- substr(taxi$pickup_datetime, 12, 13)
taxi$pickup_hour <- as.numeric(taxi$pickup_hour)
taxi$dropoff_hour <- substr(taxi$dropoff_datetime, 12, 13)
taxi$dropoff_hour <- as.numeric(taxi$dropoff_hour)
```

```{r, echo=FALSE}
# Only consider longitudes and latitudes in Manhatten
MAXLAT <- 40.85
MINLAT <- 40.675
MAXLONG <- -73.85
MINLONG <- -74.1
taxi <- filter(taxi, pickup_longitude < MAXLONG, pickup_longitude > MINLONG, 
               pickup_latitude < MAXLAT, pickup_latitude > MINLAT)
taxi <- filter(taxi, dropoff_longitude < MAXLONG, dropoff_longitude > MINLONG, 
               dropoff_latitude < MAXLAT, dropoff_latitude > MINLAT)
```

```{r, echo=FALSE}
# Remove `trip_distance` = 0
taxi <- taxi[!(taxi$trip_distance == 0),]
```

```{r, echo=FALSE}
# Remove `passenger_count` = 0 and create factor variable for passenger count
taxi <- taxi[taxi$passenger_count > 0,]
```

```{r, echo=FALSE}
# Create `trip_duration`
taxi$pickup_datetime <- ymd_hms(taxi$pickup_datetime)
taxi$dropoff_datetime <- ymd_hms(taxi$dropoff_datetime)
taxi$trip_duration <- (taxi$dropoff_datetime - taxi$pickup_datetime) / 3600
taxi$trip_duration <- as.numeric(taxi$trip_duration)
taxi$trip_duration[taxi$trip_duration == 0] <- NA
```

```{r, echo=FALSE}
# Create `avg_speed`
taxi$avg_speed <- taxi$trip_distance / taxi$trip_duration
taxi$avg_speed[taxi$avg_speed > 100] <- NA
```

```{r, echo=FALSE}
# Remove NAs
taxi <- na.omit(taxi)
```

```{r, echo=FALSE}
# Gives Heat Plot of Manhattan showing average of desired variable for each region. 
  # data: data frame from which data is drawn
  # varialbe: vector of the variable under consideration
  # lat_type: either "pickup" or "dropoff", depending on desired location by which to group data
  # GRID_RESOLUTION: integer representing number of subdivisions of longitude and latitude
  # title: string containing plot title
  # varname: string containing variable name
  # MINLAT, MAXLAT, MINLONG, MAXLONG: specify area covered by plot

locationPlot <- function(data, variable, lat_type = "pickup", GRID_RESOLUTION = 100, 
                         title = "Add Title", varname = "Variable", MINLAT = 40.675, 
                         MAXLAT = 40.85, MINLONG = -74.1, MAXLONG = -73.85) {
  
  data$variable <- variable
    
  if(lat_type == "pickup") {
    # Select all taxi rides whose pickup was within specified region. Stores 
    # these rides to a new data frame 'pos'
    pos <- filter(data, pickup_longitude < MAXLONG, pickup_longitude > MINLONG, 
                  pickup_latitude < MAXLAT, pickup_latitude > MINLAT) 
    # Remove all rows with dropoff longitude or latitude equal to 0
    pos <- filter(pos, dropoff_longitude != 0, dropoff_latitude!= 0) 
    # Remove all rows where pickup and dropoff locations are the same
    pos <- filter(pos, dropoff_longitude != pickup_longitude, 
                  dropoff_latitude != pickup_latitude) 
    # Cut latitude into bins
    pos$lat_region <- cut(pos$pickup_latitude,GRID_RESOLUTION) 
    # Cut longitude into bins
    pos$long_region <- cut(pos$pickup_longitude,GRID_RESOLUTION) 
  } 
  
  else if (lat_type == "dropoff") {
    pos <- filter(data, dropoff_longitude < MAXLONG, dropoff_longitude > MINLONG, 
                  dropoff_latitude < MAXLAT, dropoff_latitude > MINLAT) 
    pos <- filter(pos, dropoff_longitude != 0, dropoff_latitude!= 0) 
    pos <- filter(pos, dropoff_longitude != pickup_longitude, dropoff_latitude != pickup_latitude)
    pos$lat_region <- cut(pos$dropoff_latitude,GRID_RESOLUTION)
    pos$long_region <- cut(pos$dropoff_longitude,GRID_RESOLUTION)
  }
  
  # Groups data in 'pos' by 'lat_region' and 'long_region' (region on the grid) 
  # Calculates number of rides being picked up in each region ('count') and the 
  # average direction traveled for each taxi rides beginning in each grid 
  # (long_dir_avg, lat_dir_avg)
  grouped <- group_by(pos, lat_region, long_region) %>% 
    summarize(count = n(), var = mean(variable)) 

  temp <-  grouped$lat_region
  temp <- gsub("\\(", "", temp)
  temp <- gsub("\\]", "", temp)
  temp <- strsplit(temp, ",")
  temp <- unlist(temp)
  temp <- as.numeric(temp)
  grouped$max_lat <- temp[2*1:nrow(grouped)] # list of latitude upper bounds
  grouped$min_lat <- temp[2*1:nrow(grouped)-1] # listof latitude lower bounds
  grouped$avg_lat <- 0.5*(grouped$max_lat + grouped$min_lat)
  
  temp <-  grouped$long_region
  temp <- gsub("\\(", "", temp)
  temp <- gsub("\\]", "", temp)
  temp <- strsplit(temp, ",")
  temp <- unlist(temp)
  temp <- as.numeric(temp)
  grouped$max_long <- temp[2*1:nrow(grouped)] # list of longitude upper bounds
  grouped$min_long <- temp[2*1:nrow(grouped)-1] # listof longitude lower bounds
  grouped$avg_long <- 0.5*(grouped$max_long + grouped$min_long)
  
  # Load Google Map of desired region
  map <- get_map(location = c(MINLONG,MINLAT, MAXLONG, MAXLAT), source = "stamen", 
                 maptype = "watercolor") 
  
  ggmap(map) + 
    geom_rect(aes(x = min_long, y = max_long, xmin = min_long, xmax = max_long, 
                  ymin = min_lat, ymax = max_lat, fill = var), data = grouped) + 
    scale_fill_gradientn(varname, colors = c("purple","blue","green","yellow","orange","red")) + 
    labs(title = title, x = expression(italic("Longitude") ~ "(degrees)"), y = 
           expression(italic("Latitude") ~ "(degrees)"))
}
```

**What factors predict how much (as a percentage of their total fare) a person tips?**
* Do people tip more if there are more passengers in the car (perhaps due to 
social pressure)? 
```{r, fig.width = 3, fig.height= 2}
ggplot(taxi, aes(x = as.factor(passenger_count), y = tip_pct)) + geom_boxplot() + labs(title = "Tip % by # of Passengers", x = "Number of Passengers", y = "Percentage Tipped")
```
Tip percentage does not appear to change with more passengers. Average tip stays 
around the standard 20%. We will explore where larger groups tend to travel. 
For example, we predict that larger groups may travel to more touristy areas.

* Do people tip more if they get to where they are going more quickly?
```{r, fig.width = 2.5, fig.height= 2}
ggplot(taxi, aes(x = avg_speed, y = tip_pct)) + geom_smooth(method = lm) + labs(title = "Tip % by Avg Speed", x = "Average Speed (mph)", y = "Percentage Tipped")
```
There's a positive correlation between `avg_speed` and `tip_pct`, so passengers 
tip more if they get to where they are going quicker.

```{r, eval = FALSE, echo = FALSE}
# Break down into different speed groups
speeds <- taxi %>% mutate(avg_speedCat = cut(avg_speed, c(0, 10, 20, 30, 40))) %>%
  group_by(avg_speedCat) %>%
  summarize(mavg_speed = mean(avg_speed), mtip_pct = mean(tip_pct))
mspeedCat <- lm(mtip_pct ~ mavg_speed, data = speeds)

ggplot(speeds[1:5,], aes(x = mavg_speed, y = mtip_pct)) + 
  geom_point() +
  geom_smooth(method = lm) + 
  labs(title = "Percentage Tipped by Average Speed", x = 
    "Average Speed (mph)", y = "Percentage Tipped")
```

* All-subsets regression to explore which variables have the greatest effect on
`tip_pct`
```{r, echo=FALSE}
r1 = regsubsets(tip_pct ~ pickup_hour + pickup_latitude + pickup_longitude +  
                  dropoff_hour + dropoff_latitude + dropoff_longitude + 
                  passenger_count + tolls_amount + extra + vendor_id + 
                  avg_speed, data = taxi, nvmax = 11)
r1s <- summary(r1)
# r1s$which
# plot(r1s$rsq, main = "R-squareds of Models with Increasingly More Variables", 
#      ylab = "R-sqaured")

bestwhich <- r1s$which[which.min(r1s$cp),]
best_vars <- names(bestwhich[bestwhich == TRUE])
best_vars <- best_vars[-1]
mbest <- lm(tip_pct ~ pickup_hour + tolls_amount + extra + avg_speed, 
            data = taxi)
#plot(r1, scale="Cp", main = "Most Important Predictors in Different Size Models")
```

```{r}
coef(mbest)
```

* Additional Model Analysis
We will divide NYC into regions of NYC, to see if certain areas tip more than others.
Also, we will look at interaction plots to determine which variables to interact 
in the subsequent model.

* Do tips change by region of NYC (ie: more tips in the financial district)?
```{r, message=FALSE, warning=FALSE, fig.width = 3, fig.height= 3}
locationPlot(taxi[taxi$tip_pct <= 50,], taxi$tip_pct, "dropoff", title = "AvgTip by Region", varname = "TipPct")
```

In central Manhattan, most people tip around the standard 20%. As you go 
outward, there is more variability, which we will explore.

* What affects whether or not people tip more than 25%? (stepwise logistic 
regression)
```{r, echo = FALSE}
taxi$tip25 <- factor(ifelse(taxi$tip_pct >= 25, "Yes", "No"))
m25all <- glm(tip25 ~ pickup_hour + pickup_latitude + pickup_longitude +  
                dropoff_hour + dropoff_latitude + dropoff_longitude + 
                passenger_count + tolls_amount + extra + vendor_id + 
                avg_speed, data = taxi, family = binomial)
m25step <- step(m25all, direction = "backward", trace = 0)
```

```{r}
m25best <- glm(tip25 ~ dropoff_longitude + pickup_longitude + pickup_latitude + 
               tolls_amount + extra, data = taxi, family = binomial)
coef(m25best)
```

m25best tries to predict whether or not a passenger tips more than
25% using 5 variables deemed most significant from a backward stepwise 
regression. `pickup_latitude` has the biggest effect.

* What affects whether or not people tip less than 15%? (stepwise logistic 
regression)
```{r, eval=FALSE, echo=FALSE}
taxi$tip15 <- factor(ifelse(taxi$tip_pct <= 15, "Yes", "No"))
m15all <- glm(tip25 ~ pickup_hour + pickup_latitude + pickup_longitude +  
                dropoff_hour + dropoff_latitude + dropoff_longitude + 
                passenger_count + tolls_amount + extra + vendor_id + 
                avg_speed, data = taxi, family = binomial)
m15step <- step(m15all, direction = "backward")
```

```{r, eval=FALSE, echo=FALSE}
m15best <- glm(tip15 ~ pickup_latitude + tolls_amount + extra + avg_speed, 
               data = taxi, family = binomial)
summary(m15best)
```

**How do trips change with location and time of day?**
* What are the most popular pickup and drop off locations in New York? How does 
this change by time of day? Are the most popular pickup locations also the most 
popular dropoff locations?
```{r, warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
data <- taxi 
GRID_RESOLUTION <- 40
title <-  "Add Title"
varname <-  "Variable"
N_LOCATIONS <- 5

# Select all taxi rides whose pickup was within the specified region. Stores 
# these rides to a new data frame 'pos'
pos <- filter(data, pickup_longitude < MAXLONG, pickup_longitude > MINLONG, 
              pickup_latitude < MAXLAT, pickup_latitude > MINLAT) 

# Remove all rows with dropoff longitude or latitude equal to zero
pos <- filter(pos, dropoff_longitude != 0, dropoff_latitude!= 0) 
# Remove all rows where pickup and dropoff locations are the same.
pos <- filter(pos, dropoff_longitude != pickup_longitude, 
              dropoff_latitude != pickup_latitude) 

pos$lat_region <- cut(pos$pickup_latitude,GRID_RESOLUTION) # cuts latitude into bins
pos$long_region <- cut(pos$pickup_longitude,GRID_RESOLUTION) # cuts latitude into bins

# Groups data in 'pos' by 'lat_region' and 'long_region' (region on the grid) 
# Calculates number of rides being picked up in each region ('count') and the 
# average direction traveled for each taxi rides beginning in each grid (long_dir_avg, 
# lat_dir_avg)
grouped <- group_by(pos, lat_region, long_region) %>% summarize(count = n()) 

# Create new data frame 'max' containing the regions with the most pickups
max <- grouped[order(grouped$count, decreasing =TRUE),][1:N_LOCATIONS,] 

# Create data frame of all taxi rides starting fromthese region
max <- filter(pos, lat_region == max$lat_region & long_region == max$long_region) 

# Groups data by pickup_hour, latitude, and longitude and counts how many rides are in 
# each group
grouped_pickup <- group_by(max, pickup_hour, lat_region, long_region) %>% 
                    summarize(count = n()) 

# Add column identifying rides as "pickup"
grouped_pickup$type <- rep("pickup", nrow(grouped_pickup)) 
```

```{r, echo = FALSE,  results = 'hide', warning = FALSE, message = FALSE}
rm(pos, grouped, max)

# Select all taxi rides whose dropoff was within the specified region. Stores 
# these rides to a new data frame 'pos'
pos <- filter(data, dropoff_longitude < MAXLONG, dropoff_longitude > MINLONG, 
              dropoff_latitude < MAXLAT, dropoff_latitude > MINLAT) 

# Remove all rows with pickup longitude or latitude equal to zero.
pos <- filter(pos, pickup_longitude != 0, pickup_latitude!= 0) 
# Remove all rows where pickup and dropoff locations are the same.
pos <- filter(pos, dropoff_longitude != pickup_longitude, 
              dropoff_latitude != pickup_latitude) 

pos$lat_region <- cut(pos$dropoff_latitude,GRID_RESOLUTION) # cuts latitude into bins
pos$long_region <- cut(pos$dropoff_longitude,GRID_RESOLUTION) # cuts longitude into bins
      
# Group data in 'pos' by 'lat_region' and 'long_region' (i.e. by region on the 
# grid). Calculates number of rides being picked up in each region ('count') 
# and the average direction traveled for each taxi rides beginning in each grid 
# (long_dir_avg, lat_dir_avg)
grouped <- group_by(pos, lat_region, long_region) %>% summarize(count = n()) 

# Create new data frame 'max' containing the regions with the most dropoffs
max <- grouped[order(grouped$count, decreasing =TRUE),][1:N_LOCATIONS,]

# Create data frame of all taxi rides going to these region
max <- filter(pos, lat_region == max$lat_region & long_region == max$long_region) 

# Groups data by pickup_hour, latitude, and longitude and counts how many rides are in each group
grouped_dropoff <- group_by(max, pickup_hour, lat_region, long_region) %>% summarize(count = n()) 

# Identifies data as "dropoff"
grouped_dropoff$type <- rep("dropoff", nrow(grouped_dropoff)) 
```
      
      
```{r, echo = FALSE,  results = 'hide', warning = FALSE, message = FALSE}
# Combine data tables "grouped_pickup" and "grouped_dropoff" into  a new data 
# frame 'dat'
dat <- bind_rows(ungroup(grouped_pickup), ungroup(grouped_dropoff)) 

# Create data frame of latitudes and longitudes
grid <- data.frame(latitude = dat$lat_region,longitude = dat$long_region) 
grid <- unique(grid) # removes duplicates

grid$regioncode <- 1:nrow(grid) # gives each region its own integer code

# Classify rides in 'dat' by region
dat$region <- rep(NA,nrow(dat))
for(i in 1: nrow(grid)) {
  dat$region[(as.character(dat$lat_region) == as.character(grid$latitude[i])) & 
       (as.character(dat$long_region) == as.character(grid$longitude[i]))] <- i
} 

dat$region <- as.factor(dat$region)
```

```{r, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
temp <-  grid$latitude
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")

# Split region boundaries from 'cut()' into upper and lower latitudes
temp <- unlist(temp)
temp <- as.numeric(temp)
grid$max_lat <- temp[2*1:nrow(grid)] # list of latitude upper bounds
grid$min_lat <- temp[2*1:nrow(grid)-1] # list of latitude lower bounds
grid$avg_lat <- 0.5*(grid$max_lat + grid$min_lat) # list of average region latitudes
    
temp <-  grid$longitude # stores longitude bounds of regions to 'temp'
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")

# Split region boundaries from 'cut()' into upper and lower longitudes
temp <- unlist(temp)
temp <- as.numeric(temp)
grid$max_long <- temp[2*1:nrow(grid)] # list of latitude upper bounds
grid$min_long <- temp[2*1:nrow(grid)-1] # list of latitude lower bounds
grid$avg_long <- 0.5*(grid$max_long + grid$min_long) # list of average region longitudes
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
timeplot <- ggplot(data = dat, aes(x = pickup_hour, y = count, col = region)) +
  geom_line(size = 1) +
  labs(title = "Analysis of 5 Most Popular Pickup and Dropoff Sites", 
       x = expression(italic("Hour of Day")), 
       y = expression(italic("Number of Rides"))) + 
  facet_wrap(~ type, ncol =1)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
MAXLAT1 <- 40.775
MINLAT1 <- 40.74
MAXLONG1 <- -73.965
MINLONG1<- -73.995
map <- get_map(location = c(MINLONG1 - 0.01 ,MINLAT1 - 0.01, MAXLONG1 + 0.01, 
                     MAXLAT1 + 0.01), source = "google", maptype = "roadmap") 

pop_map <- ggmap(map) +
  labs(x = expression(italic("Longitude")~"(degrees)"),
       y = expression(italic("Latitude")~"(degrees)")) +
  geom_label(aes(jitter(avg_long), jitter(avg_lat),label = regioncode), 
             data = grid, size = 2) + 
  theme(axis.line = element_blank(), 
    axis.text.x = element_blank(), 
    axis.text.y = element_blank(),
    axis.ticks = element_blank(), 
    axis.title.x = element_blank(), 
    axis.title.y = element_blank(),plot.margin=unit(c(0,0,0,0),"mm")) +
  scale_y_continuous(limits=c(MINLAT1, MAXLAT1)) + 
  scale_x_continuous(limits = c(MINLONG1, MAXLONG1))
```

```{r, warning = FALSE, message = FALSE, fig.width = 5, fig.height= 3}
grid.arrange(timeplot, pop_map, ncol = 2, widths = c(2,1))
```
Regions 1-5 are the most popular taxi pickup sites (1 = most popular, 5 = least). 
egions 6-10 are the most popular dropoff sites. Dropoff sites tend be most 
active at 7-8 AM. Pickup sites are most active at 8-9 PM. We will next 
investigate how popular pickup sites rank in terms of dropoff sites and vice 
versa.
  
* Which directions are taxis traveling on average at different times of the day? 
```{r, warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
# Remove rows with abnormal dropoff longitudes
rm(pos, grouped)
MAXLAT <- 40.85
MINLAT <- 40.675
MAXLONG <- -73.85
MINLONG <- -74.1
SIZEFACTOR  <-  0.015 # Desired Magnitude of Direction Arrows
GRID_RESOLUTION <- 25

# Select all taxi rides whose pickup was within the specified region. Stores 
# these rides to a new data frame 'pos'
pos <- filter(taxi, pickup_longitude < MAXLONG, pickup_longitude > MINLONG, 
              pickup_latitude < MAXLAT, pickup_latitude > MINLAT) 

# Remove all rows with dropoff longitude or latitude equal to zero.
pos <- filter(pos, dropoff_longitude != 0, dropoff_latitude!= 0) 

# Remove all rows where pickup and dropoff locations are the same.
pos <- filter(pos, dropoff_longitude != pickup_longitude, 
              dropoff_latitude != pickup_latitude) 

pos$lat_region <- cut(pos$pickup_latitude,GRID_RESOLUTION) # cuts latitude into bins
pos$long_region <- cut(pos$pickup_longitude,GRID_RESOLUTION) # cuts latitude into bin
```

```{r, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
# Add rows 'lat_dir' and 'long_dir' to the data frame 'pos' which respectively 
# contain the change in latitude and longitude for each ride.
pos <- mutate(pos, lat_dir = (dropoff_latitude - pickup_latitude), 
              long_dir = (dropoff_longitude - pickup_longitude)) 

# Create columns 'lat_dir_scaled' and 'long_dir_scaled', which contain the 
# coordinates of 'lat_dir' and 'long_dir' normalized so that the direction 
# vector has magnitude equal to 'SIZEFACTOR'
pos <- mutate(pos, lat_dir_scaled = lat_dir * SIZEFACTOR/sqrt(lat_dir^2 + long_dir^2), 
              long_dir_scaled = long_dir * SIZEFACTOR /sqrt(lat_dir^2 + long_dir^2)) 

# Note: NaN results for rides where pickup and dropoff locations are equal.

pos$daytime <- rep(NA, nrow(pos))

# Create new column of pos indicating whether pickup-time was in the morning or evening
pos$daytime <- ifelse(pos$pickup_hour < 12, "morning", "evening") 
pos$daytime <- factor(pos$daytime, levels= c("morning","evening"))
```

```{r,warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
# Groups data in 'pos' by 'lat_region' and 'long_region' (i.e. by region on the 
# grid). Calculates number of rides being picked up in each region ('count') and 
# the average direction traveled for each taxi rides beginning in each grid 
# (long_dir_avg, lat_dir_avg)
grouped <- group_by(pos, lat_region, long_region, daytime) %>% 
  summarize(lat_dir_avg = mean(lat_dir_scaled, na.rm = TRUE), 
            long_dir_avg = mean(long_dir_scaled, na.rm =TRUE), count = n()) 

temp <-  grouped$lat_region 
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")
temp <- unlist(temp)
temp <- as.numeric(temp)
grouped$max_lat <- temp[2*1:nrow(grouped)] # list of latitude upper bounds
grouped$min_lat <- temp[2*1:nrow(grouped)-1] # list of latitude lower bounds
grouped$avg_lat <- 0.5*(grouped$max_lat + grouped$min_lat) # list of average latitude
    
temp <-  grouped$long_region
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")
temp <- unlist(temp)
temp <- as.numeric(temp)
grouped$max_long <- temp[2*1:nrow(grouped)] # list of latitude upper bounds
grouped$min_long <- temp[2*1:nrow(grouped)-1] # list of latitude lower bounds
grouped$avg_long <- 0.5*(grouped$max_long + grouped$min_long) # list of average longitude
```

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
heatplot <- ggplot(data = grouped, aes(xmin = min_long, xmax = max_long, 
                   ymin = min_lat, ymax = max_lat, fill = count)) + 
  geom_rect() + 
  scale_fill_gradientn("Number of Pickups", 
                       colors = c("purple","blue","green","yellow", "orange", "red")) + 
  facet_wrap(~daytime)
```

```{r, warning = FALSE, message = FALSE, echo=FALSE, fig.width = 9, fig.height= 3}
MAXLAT1 <- 40.8
MINLAT1 <- 40.7
MAXLONG1 <- -73.92
MINLONG1<- -74.03
MINRIDES <-  30 # minimum number of rides for arrow to show up

map <- get_map(location = c(MINLONG, MINLAT, MAXLONG, MAXLAT), source = "stamen", 
               maptype = "watercolor", color = "bw")
```

```{r, warning = FALSE, message = FALSE, fig.width = 5.5, fig.height= 3}
ggmap(map) + geom_segment(aes(x = avg_long, xend = (avg_long + long_dir_avg), y = avg_lat, yend = (avg_lat + lat_dir_avg), color = count), data = filter(grouped, count > MINRIDES), arrow = arrow(length=unit(0.15, "cm"), ends="first", type = "closed"), size = 0.9) + labs(title = "Direction of Motion by Region", x = expression(italic("Longitude")~"(degrees)"), y = expression(italic("Latitude")~"(degrees)")) + scale_color_gradientn("Number of Rides", colors = c("purple","blue","green","yellow","orange","red")) + facet_wrap( ~ daytime) + scale_y_continuous(limits=c(MINLAT1, MAXLAT1)) + scale_x_continuous(limits = c(MINLONG1, MAXLONG1))
```
There is a general outward flux of taxis. Taxi rides starting farther from 
center tend to agree more on the outward direction (longer arrows). We will 
facet into more time catagories. We might investigate where outward directed 
taxis are going

* What is the average speed of taxis depending on the time of day and region of 
the city?
To do this, we would create a heat map of Manhattan where different colors 
indicate different travel speeds of taxis.

