---
title: "Taxi Final Report"
author: "Christoph, Chris, Carol"
date: "4/30/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 2015 Yellow Taxi Trip Data - NYC Open Data
https://data.cityofnewyork.us/Transportation/2015-Yellow-Taxi-Trip-Data/ba8s-jw6u

"This dataset includes trip records from all trips completed in yellow taxis 
in NYC from January to June in 2015.  Records include fields capturing pick-up 
and drop-off dates/times, pick-up and drop-off locations, trip distances, 
itemized fares, rate types, payment types, and driver-reported passenger counts. 
The data used in the attached datasets were collected and provided to the NYC 
Taxi and Limousine Commission (TLC) by technology providers authorized under the 
Taxicab Passenger Enhancement Program (TPEP). The trip data was not created by 
the TLC, and TLC makes no representations as to the accuracy of these data." 
(from NYC Open Data)

This dataset is huge (146,087,462 rows and 5GB), and too big for our computers 
to work with, so we decided to load in data for 1 day. We chose May 6, for no 
reason in partiocular, except that it is a non-holiday weekday. This was still
a huge amount of data however, so we filtered it further by selecting only the 
rides where passengers paid with a credit card (because these are the only rides
with tip data, which we are interested in exploring), and then took a random 10%
of the rides.

## Variables
* `avg_speed`: average speed of the taxi during the trip in mph
* `dropoff_datetime`: date and time when meter was disengaged
* `dropoff_hour`: hour of the day when the meter was disengaged
* `dropoff_latitude`: latitude where the meter was disengaged
* `dropoff_longitude`: longitude where the meter was disengaged
* `dropoff_region`: region where the meter was disenaged (one of 25 regions into 
                    which we divided NYC)
* `dropoff_time`: time of day when meter was disengaged (either morning, 
                  afternoon, evening, or night)
* `extra`: Miscellaneous extras and surcharges (0 = no charge, 0.5 = $0.50 rush 
           hour charge, 1 = $1 overnight charge)
* `fare_amount`: time-and-distance fare calculated by the meter
* `mta_tax`: $0.50 tax on taxi rides        
* `passenger_count`: number of passengers in the vehicle
* `payment_type`: how the passenger paid for the trip. Note: Always 1 (credit 
                  card) in the subset of the data we are analyzing. Only about 
                  1/3 of the data was not 1
* `pickup_datetime`: date and time when meter was engaged 
* `pickup_hour`: hour of the day when the meter was engaged
* `pickup_latitude`: latitude where the meter was engaged   
* `pickup_longitude`: longitude where the meter was engaged   
* `pickup_region`: region where the meter was enaged (one of 25 regions into 
                    which we divided NYC)
* `pickup_time`: time of day when meter was engaged (either morning, afternoon,
                 evening, or night)
* `rate_code`: final rate code in effect at the end of the trip (1 = Standard 
               rate, 2 = JFK, 3 = Newark 4 = Nassau or Westchester, 5 = 
               Negotiated Fare, 6 = Group Ride)       
* `store_and_fwd_flag`: whether the trip record was held in vehicle memory 
                        before sending to the vendor because the vehicle did not 
                        have a connection to the server (Y = yes, N = no)
* `tip_amount`: tip paid. Note: Only calculated for credit card tips   
* `tip_pct` - tip as a percentage of total fee (fare + tolls + extra + mta_tax)
* `tolls_amount`: total amount of all tolls paid in the trip
* `total_amount`: total amount charged to passengers (fare + tip + tolls + 
                  extra + mta_tax)
* `trip_distance`: elapsed trip distance in miles
* `trip_duration`: length of time, in minutes, that the meter was engaged
* `vendor_id`: technology vendor that provided the record (1 = Creative Mobile 
               Technologies, 2 = VeriFone)

## Research Questions
We are particularly interested in exploring 3 features of this taxi trips 
dataset: time (`dropoff_datetime` and `pickup_datetime`), location 
(`dropoff_latitude`, `dropoff_longitude`, `pickup_latitude`, and
`pickup_longitude`), and payments, specifically tips (`tip_amount`). 

  * What factors predict how much (as a percentage of their total fare) a person 
    tips?
  * What are the most popular pickup and drop off locations in New York and at 
    what times are they most popular? 
    * Which directions are taxis traveling at different times of the day? 
  
```{r, echo=FALSE}
taxi <- read.csv("taxi_data.csv", as.is = TRUE)
```

```{r, echo = FALSE, meassage = FALSE, warning = FALSE, include = FALSE}
library(dplyr)
library(ggplot2)
library(leaps)
library(lubridate)
library(MASS)
library(ggmap)
library(mapproj)
library(gridExtra)
library(grid)
```


```{r, echo=FALSE}
# Create `tip_pct`
taxi <- taxi %>% mutate(tip_pct = (tip_amount / (total_amount - tip_amount)) * 100)

# Remove observation where tip_pct > 100
taxi <- taxi[taxi$tip_pct <= 100,]
```


```{r, echo=FALSE}
# Create `pickup_hour` and `dropoff_hour`
taxi$pickup_hour <- as.numeric(substr(taxi$pickup_datetime, 12, 13))
taxi$dropoff_hour <- as.numeric(substr(taxi$dropoff_datetime, 12, 13))

# Create `pickup_time` and `dropoff_time`
taxi$pickup_time <- factor(ifelse(taxi$pickup_hour %in% 6:11, "Morning", 
                    ifelse(taxi$pickup_hour %in% 12:17, "Afternoon", 
                    ifelse(taxi$pickup_hour %in% 18:23, "Evening",
                    ifelse(taxi$pickup_hour %in% 0:5, "Night", NA)))), 
                    levels = c("Morning", "Afternoon", "Evening", "Night"))
taxi$dropoff_time <- factor(ifelse(taxi$dropoff_hour %in% 6:11, "Morning", 
                     ifelse(taxi$dropoff_hour %in% 12:17, "Afternoon", 
                     ifelse(taxi$dropoff_hour %in% 18:23, "Evening",
                     ifelse(taxi$dropoff_hour %in% 0:5, "Night", NA)))),
                     levels = c("Morning", "Afternoon", "Evening", "Night"))
```


```{r, echo=FALSE}
# Only consider longitudes and latitudes in Manhatten
MAXLAT <- 40.85; MINLAT <- 40.675; MAXLONG <- -73.85; MINLONG <- -74.1
taxi <- filter(taxi, pickup_longitude < MAXLONG, pickup_longitude > MINLONG, 
               pickup_latitude < MAXLAT, pickup_latitude > MINLAT)
taxi <- filter(taxi, dropoff_longitude < MAXLONG, dropoff_longitude > MINLONG, 
               dropoff_latitude < MAXLAT, dropoff_latitude > MINLAT)

# Remove rows where longitude/latitude = 0 or dropoff = pickup
taxi <- filter(taxi, dropoff_longitude != 0, dropoff_latitude != 0) 
taxi <- filter(taxi, dropoff_longitude != pickup_longitude, 
              dropoff_latitude != pickup_latitude) 
```

```{r, echo=FALSE}
# Create geographic regions 
taxi$dropoff_lat_region <- cut(taxi$dropoff_latitude, 5)
taxi$dropoff_long_region <- cut(taxi$dropoff_longitude, 5)
taxi$pickup_lat_region <- cut(taxi$pickup_latitude, 5)
taxi$pickup_long_region <- cut(taxi$pickup_longitude, 5)

# pickup_region <- taxi %>% group_by(pickup_lat_region, pickup_long_region) %>% 
#   summarize(count = n(), var = mean(tip_pct)) 
# dropoff_region <- taxi %>% group_by(dropoff_lat_region, dropoff_long_region) %>% 
#   summarize(count = n(), var = mean(tip_pct))

# taxi$pickup_region <- rep(0, nrow(taxi))
# for(i in 1:nrow(taxi)) {
#   count <- 1
#   for(j in levels(taxi$pickup_lat_region)) {
#     for(k in levels(taxi$pickup_long_region)) {
#       if(taxi$pickup_lat_region[i] == j & taxi$pickup_long_region[i] == k) {
#         taxi$pickup_region[i] <- count
#       }
#       count <- count + 1
#     }
#   }
# }
# 
# taxi$dropoff_region <- rep(0, nrow(taxi))
# for(i in 1:nrow(taxi)) {
#   count <- 1
#   for(j in levels(taxi$dropoff_lat_region)) {
#     for(k in levels(taxi$dropoff_long_region)) {
#       if(taxi$dropoff_lat_region[i] == j && taxi$dropoff_long_region[i] == k) {
#         taxi$dropoff_region[i] <- count
#       }
#       count <- count + 1
#     }
#   }
# }
```


```{r, echo=FALSE}
# Remove `trip_distance` = 0
taxi <- taxi[!(taxi$trip_distance == 0),]
```


```{r, echo=FALSE}
# Remove `passenger_count` = 0 and create factor variable for passenger count
taxi <- taxi[taxi$passenger_count > 0,]
```


```{r, echo=FALSE}
# Create `trip_duration`
taxi$pickup_datetime <- ymd_hms(taxi$pickup_datetime)
taxi$dropoff_datetime <- ymd_hms(taxi$dropoff_datetime)
taxi$trip_duration <- as.numeric((taxi$dropoff_datetime - taxi$pickup_datetime) / 60)
taxi$trip_duration[taxi$trip_duration == 0] <- 1
```


```{r, echo=FALSE}
# Create `avg_speed`
taxi$avg_speed <- taxi$trip_distance / (taxi$trip_duration / 60)
taxi$avg_speed[taxi$avg_speed > 50] <- 50
```


```{r, echo=FALSE}
# Remove NAs
which(is.na(taxi))
taxi <- na.omit(taxi)
```


```{r, echo=FALSE}
# Produces heat plot of NYC showing average of desired variable for each region. 
  # data: data frame from which data is drawn
  # variable: vector of the variable under consideration
  # lat_type: either "pickup" or "dropoff"
  # GRID_RESOLUTION: integer representing number of divisions of longitude/latitude
  # title: string containing plot title
  # varname: string containing variable name
  # MINLAT, MAXLAT, MINLONG, MAXLONG: specify area covered by plot

locationPlot <- function(data, variable, lat_type = "pickup", GRID_RESOLUTION = 100, 
                         title = "Add Title", varname = "Variable", MINLAT = 40.675, 
                         MAXLAT = 40.85, MINLONG = -74.1, MAXLONG = -73.85) {
  
  data$variable <- variable
    
  if(lat_type == "pickup") {
    # Select all taxi rides whose pickup was within specified region. Stores 
    # these rides to a new data frame 'pos'
    pos <- filter(data, pickup_longitude < MAXLONG, pickup_longitude > MINLONG, 
                  pickup_latitude < MAXLAT, pickup_latitude > MINLAT) 
    # Remove all rows with dropoff longitude or latitude equal to 0
    pos <- filter(pos, dropoff_longitude != 0, dropoff_latitude!= 0) 
    # Remove all rows where pickup and dropoff locations are the same
    pos <- filter(pos, dropoff_longitude != pickup_longitude, 
                  dropoff_latitude != pickup_latitude) 
    # Cut latitude into bins
    pos$lat_region <- cut(pos$pickup_latitude,GRID_RESOLUTION) 
    # Cut longitude into bins
    pos$long_region <- cut(pos$pickup_longitude,GRID_RESOLUTION) 
  } 
  
  else if (lat_type == "dropoff") {
    pos <- filter(data, dropoff_longitude < MAXLONG, dropoff_longitude > MINLONG, 
                  dropoff_latitude < MAXLAT, dropoff_latitude > MINLAT) 
    pos <- filter(pos, dropoff_longitude != 0, dropoff_latitude!= 0) 
    pos <- filter(pos, dropoff_longitude != pickup_longitude, dropoff_latitude != pickup_latitude)
    pos$lat_region <- cut(pos$dropoff_latitude,GRID_RESOLUTION)
    pos$long_region <- cut(pos$dropoff_longitude,GRID_RESOLUTION)
  }
  
  # Groups data in 'pos' by 'lat_region' and 'long_region' (region on the grid) 
  # Calculates number of rides being picked up in each region ('count') and the 
  # average direction traveled for each taxi rides beginning in each grid 
  # (long_dir_avg, lat_dir_avg)
  grouped <- group_by(pos, lat_region, long_region) %>% 
    summarize(count = n(), var = mean(variable)) 

  temp <-  grouped$lat_region
  temp <- gsub("\\(", "", temp)
  temp <- gsub("\\]", "", temp)
  temp <- strsplit(temp, ",")
  temp <- unlist(temp)
  temp <- as.numeric(temp)
  grouped$max_lat <- temp[2*1:nrow(grouped)] # list of latitude upper bounds
  grouped$min_lat <- temp[2*1:nrow(grouped)-1] # listof latitude lower bounds
  grouped$avg_lat <- 0.5*(grouped$max_lat + grouped$min_lat)
  
  temp <-  grouped$long_region
  temp <- gsub("\\(", "", temp)
  temp <- gsub("\\]", "", temp)
  temp <- strsplit(temp, ",")
  temp <- unlist(temp)
  temp <- as.numeric(temp)
  grouped$max_long <- temp[2*1:nrow(grouped)] # list of longitude upper bounds
  grouped$min_long <- temp[2*1:nrow(grouped)-1] # listof longitude lower bounds
  grouped$avg_long <- 0.5*(grouped$max_long + grouped$min_long)
  
  # Load Google Map of desired region
  map <- get_map(location = c(MINLONG,MINLAT, MAXLONG, MAXLAT), source = "stamen", 
                 maptype = "watercolor") 
  
  ggmap(map) + 
    geom_rect(aes(x = min_long, y = max_long, xmin = min_long, xmax = max_long, 
                  ymin = min_lat, ymax = max_lat, fill = var), data = grouped) + 
    scale_fill_gradientn(varname, colors = c("purple","blue","green","yellow","orange","red")) + 
    labs(title = title, x = expression(italic("Longitude") ~ "(degrees)"), y = 
           expression(italic("Latitude") ~ "(degrees)"))
}
```

**What factors predict how much (as a percentage of their total fare) a person tips?**

* Do people tip more if there are more passengers in the car (perhaps due to 
social pressure)? 

```{r, fig.width = 3, fig.height= 2}
ggplot(taxi, aes(x = as.factor(passenger_count), y = tip_pct)) + 
  geom_boxplot() + 
  labs(title = "Tip % by # of Passengers", x = "Number of Passengers", 
       y = "Percentage Tipped")
```

```{r, echo=FALSE}
mpassengers <- lm(tip_pct ~ passenger_count, data = taxi)
summary(mpassengers)
anova(mpassengers)
pairwise.t.test(taxi$tip_pct, taxi$passenger_count)
```

Tip percentage does not appear to change with more passengers. Average tip stays 
around the standard 20%.


* Do people tip more if they get to where they are going more quickly?

```{r, fig.width = 2.5, fig.height= 2}
ggplot(subset(taxi, tip_pct <= 50), aes(x = avg_speed, y = tip_pct)) + 
  geom_point() +
  geom_smooth() + 
  labs(title = "Tip % by Avg Speed", x = "Average Speed (mph)", 
       y = "Percentage Tipped")
```

```{r, echo=FALSE}
mspeed <- lm(tip_pct ~ avg_speed, data = taxi)
summary(mspeed)
```

```{r, eval = FALSE, echo = FALSE}
# Break down into different speed groups
speeds <- taxi %>% mutate(avg_speedCat = cut(avg_speed, c(0, 10, 20, 30, 40))) %>%
  group_by(avg_speedCat) %>%
  summarize(mavg_speed = mean(avg_speed), mtip_pct = mean(tip_pct))

mspeedCat <- lm(mtip_pct ~ mavg_speed, data = speeds)
summary(mspeedCat)

ggplot(speeds[1:5,], aes(x = mavg_speed, y = mtip_pct)) + 
  geom_point() +
  geom_smooth(method = lm) + 
  labs(title = "Percentage Tipped by Average Speed", x = 
    "Average Speed (mph)", y = "Percentage Tipped")
```


* Do people tip more, as a percentage, if their fare is lower?

```{r, echo=FALSE}
mfare <- lm(tip_pct ~ fare_amount, data = taxi)
summary(mfare)

ggplot(subset(taxi, fare_amount < 50), aes(x = fare_amount, y = tip_pct)) + 
  geom_point() + 
  geom_smooth() +
  labs(title = "Percentage Tipped by Fare Amount", x = "Fare Amount ($)", 
       y = "Percentage Tipped")
```

```{r, echo=FALSE}
# Including tolls and extra charges
mtotal <- lm(tip_pct ~ fare_amount + tolls_amount + extra + mta_tax, data = taxi)
summary(mtotal)

ggplot(subset(taxi, fare_amount < 50), aes(x = total_amount - tip_amount, y = tip_pct)) + 
  geom_point() + 
  geom_smooth() +
  labs(title = "Percentage Tipped by Total Amount", x = "Total Amount ($)", 
       y = "Percentage Tipped")
```


* Do people tip more at different times of day?

```{r, echo=FALSE}
mpickup_time <- lm(tip_pct ~ pickup_time, data = taxi)
summary(mpickup_time)

mdropoff_time <- lm(tip_pct ~ dropoff_time, data = taxi)
summary(mdropoff_time)

mtime <- lm(tip_pct ~ pickup_time + dropoff_time, data = taxi)
summary(mtime)
```

```{r, echo=FALSE}
ggplot(taxi, aes(x = pickup_time, y = tip_pct)) + 
  geom_boxplot() + 
  labs(title = "Percentage Tipped by Time of Day", x = "Time of Day", y = "Percentage Tipped")
```

```{r, echo=FALSE}
ggplot(taxi, aes(x = pickup_hour, y = tip_pct)) + 
  geom_jitter() + 
  labs(title = "Percentage Tipped by Time of Day", x = "Time of Day", y = "Percentage Tipped")
```


* Do people tip more in different areas?

```{r, echo=FALSE}
mpickup_lat <- lm(tip_pct ~ pickup_lat_region, data = taxi)
summary(mpickup_lat)
mpickup_long <- lm(tip_pct ~ pickup_long_region, data = taxi)
summary(mpickup_long)
mdropoff_lat <- lm(tip_pct ~ dropoff_lat_region, data = taxi)
summary(mdropoff_lat)
mdropoff_long <- lm(tip_pct ~ dropoff_long_region, data = taxi)
summary(mdropoff_long)

mlat <- lm(tip_pct ~ pickup_lat_region + dropoff_lat_region, data = taxi)
summary(mlat)
mlong <- lm(tip_pct ~ pickup_long_region + dropoff_long_region, data = taxi)
summary(mlong)

mregion <- lm(tip_pct ~ factor(pickup_region) + factor(dropoff_region), 
              data = taxi)
summary(mregion)
```

```{r, message=FALSE, warning=FALSE, fig.width = 3, fig.height= 3}
locationPlot(taxi, taxi$tip_pct, lat_type = "pickup", GRID_RESOLUTION = 30, title = 
               "Percentage Tipped by Time of Day", varname = "TipPct")

locationPlot(taxi, taxi$tip_pct, lat_type = "dropoff", GRID_RESOLUTION = 30, title = 
               "Percentage Tipped by Time of Day", varname = "TipPct")
```


* Which variables are most important for predicting `tip_pct`. All-subsets 
regression 

```{r, echo=FALSE}
r1 = regsubsets(tip_pct ~ extra + fare_amount + mta_tax + passenger_count + 
                  tolls_amount + trip_distance + trip_duration + avg_speed + 
                  pickup_time + dropoff_time + pickup_region + dropoff_region, 
                  data = taxi, nvmax = 12)
r1s <- summary(r1)
r1s$which
plot(r1s$rsq, main = "R-squareds of Models with Increasingly More Variables", 
     ylab = "R-sqaured")

bestwhich <- r1s$which[which.min(r1s$cp),]
best_vars <- names(bestwhich[bestwhich == TRUE])
best_vars <- best_vars[-1]
mbest <- lm(tip_pct ~ fare_amount + tolls_amount + trip_distance + trip_duration +
            avg_speed + dropoff_region, data = taxi)
summary(mbest)
coef(mbest)
plot(r1, scale="Cp", main = "Most Important Predictors in Different Size Models")
```


* What affects whether or not people tip more than 25%? (stepwise logistic 
regression)

```{r, echo = FALSE}
taxi$tip25 <- factor(ifelse(taxi$tip_pct >= 25, "Yes", "No"))
m25all <- glm(tip25 ~ extra + fare_amount + mta_tax + passenger_count + 
              tolls_amount + trip_distance + trip_duration + avg_speed + 
              pickup_time + dropoff_time + pickup_region + dropoff_region, 
              data = taxi, family = binomial)
m25step <- step(m25all, direction = "backward", trace = 0)
summary(m25step)
coef(m25step)
```

m25best tries to predict whether or not a passenger tips more than
25% using 5 variables deemed most significant from a backward stepwise 
regression. `pickup_latitude` has the biggest effect.


* What affects whether or not people tip less than 15%? (stepwise logistic 
regression)

```{r, echo=FALSE}
taxi$tip15 <- factor(ifelse(taxi$tip_pct <= 15, "Yes", "No"))
m15all <- glm(tip15 ~ extra + fare_amount + mta_tax + passenger_count + 
              tolls_amount + trip_distance + trip_duration + avg_speed + 
              pickup_time + dropoff_time + pickup_region + dropoff_region, 
              data = taxi, family = binomial)
m15step <- step(m15all, direction = "backward")
summary(m15step)
coef(m15step)
```

* What affects whether or not people tip at all? (stepwise logistic regression)

```{r, echo=FALSE}
taxi$tip0 <- factor(ifelse(taxi$tip_pct == 0, "Yes", "No"))
m0all <- glm(tip0 ~ extra + fare_amount + mta_tax + passenger_count + 
              tolls_amount + trip_distance + trip_duration + avg_speed + 
              pickup_time + dropoff_time + pickup_region + dropoff_region, 
              data = taxi, family = binomial)
m0step <- step(m0all, direction = "backward")
summary(m0step)
coef(m0step)
```


**How do trips change with location and time of day?**

* What are the most popular pickup and drop off locations in New York? How does 
this change by time of day? Are the most popular pickup locations also the most 
popular dropoff locations?

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
data <- taxi 
GRID_RESOLUTION <- 40
title <-  "Add Title"
varname <-  "Variable"
N_LOCATIONS <- 5

# Select all taxi rides whose pickup was within the specified region. Stores 
# these rides to a new data frame 'pos'
pos <- filter(data, pickup_longitude < MAXLONG, pickup_longitude > MINLONG, 
              pickup_latitude < MAXLAT, pickup_latitude > MINLAT) 

# Remove all rows with dropoff longitude or latitude equal to zero
pos <- filter(pos, dropoff_longitude != 0, dropoff_latitude!= 0) 
# Remove all rows where pickup and dropoff locations are the same.
pos <- filter(pos, dropoff_longitude != pickup_longitude, 
              dropoff_latitude != pickup_latitude) 

pos$lat_region <- cut(pos$pickup_latitude,GRID_RESOLUTION) # cuts latitude into bins
pos$long_region <- cut(pos$pickup_longitude,GRID_RESOLUTION) # cuts latitude into bins

# Groups data in 'pos' by 'lat_region' and 'long_region' (region on the grid) 
# Calculates number of rides being picked up in each region ('count') and the 
# average direction traveled for each taxi rides beginning in each grid (long_dir_avg, 
# lat_dir_avg)
grouped <- group_by(pos, lat_region, long_region) %>% summarize(count = n()) 

# Create new data frame 'max' containing the regions with the most pickups
max <- grouped[order(grouped$count, decreasing =TRUE),][1:N_LOCATIONS,] 

# Create data frame of all taxi rides starting from these regions
max <- filter(pos, lat_region == max$lat_region & long_region == max$long_region) 

# Groups data by pickup_hour, latitude, and longitude and counts how many rides are in 
# each group
grouped_pickup <- group_by(max, pickup_hour, lat_region, long_region) %>% 
                    summarize(count = n()) 

# Add column identifying rides as "pickup"
grouped_pickup$type <- rep("pickup", nrow(grouped_pickup)) 
```

```{r, echo = FALSE,  results = 'hide', warning = FALSE, message = FALSE}
rm(pos, grouped, max)

# Select all taxi rides whose dropoff was within the specified region. Stores 
# these rides to a new data frame 'pos'
pos <- filter(data, dropoff_longitude < MAXLONG, dropoff_longitude > MINLONG, 
              dropoff_latitude < MAXLAT, dropoff_latitude > MINLAT) 

# Remove all rows with pickup longitude or latitude equal to zero.
pos <- filter(pos, pickup_longitude != 0, pickup_latitude!= 0) 
# Remove all rows where pickup and dropoff locations are the same.
pos <- filter(pos, dropoff_longitude != pickup_longitude, 
              dropoff_latitude != pickup_latitude) 

pos$lat_region <- cut(pos$dropoff_latitude,GRID_RESOLUTION) # cuts latitude into bins
pos$long_region <- cut(pos$dropoff_longitude,GRID_RESOLUTION) # cuts longitude into bins
      
# Group data in 'pos' by 'lat_region' and 'long_region' (i.e. by region on the 
# grid). Calculates number of rides being picked up in each region ('count') 
# and the average direction traveled for each taxi rides beginning in each grid 
# (long_dir_avg, lat_dir_avg)
grouped <- group_by(pos, lat_region, long_region) %>% summarize(count = n()) 

# Create new data frame 'max' containing the regions with the most dropoffs
max <- grouped[order(grouped$count, decreasing =TRUE),][1:N_LOCATIONS,]

# Create data frame of all taxi rides going to these region
max <- filter(pos, lat_region == max$lat_region & long_region == max$long_region) 

# Groups data by pickup_hour, latitude, and longitude and counts how many rides are in each group
grouped_dropoff <- group_by(max, pickup_hour, lat_region, long_region) %>% summarize(count = n()) 

# Identifies data as "dropoff"
grouped_dropoff$type <- rep("dropoff", nrow(grouped_dropoff)) 
```
      
      
```{r, echo = FALSE,  results = 'hide', warning = FALSE, message = FALSE}
# Combine data tables "grouped_pickup" and "grouped_dropoff" into  a new data 
# frame 'dat'
dat <- bind_rows(ungroup(grouped_pickup), ungroup(grouped_dropoff)) 

# Create data frame of latitudes and longitudes
grid <- data.frame(latitude = dat$lat_region,longitude = dat$long_region) 
grid <- unique(grid) # removes duplicates

grid$regioncode <- 1:nrow(grid) # gives each region its own integer code

# Classify rides in 'dat' by region
dat$region <- rep(NA,nrow(dat))
for(i in 1: nrow(grid)) {
  dat$region[(as.character(dat$lat_region) == as.character(grid$latitude[i])) & 
       (as.character(dat$long_region) == as.character(grid$longitude[i]))] <- i
} 

dat$region <- as.factor(dat$region)
```

```{r, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
temp <-  grid$latitude
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")

# Split region boundaries from 'cut()' into upper and lower latitudes
temp <- unlist(temp)
temp <- as.numeric(temp)
grid$max_lat <- temp[2*1:nrow(grid)] # list of latitude upper bounds
grid$min_lat <- temp[2*1:nrow(grid)-1] # list of latitude lower bounds
grid$avg_lat <- 0.5*(grid$max_lat + grid$min_lat) # list of average region latitudes
    
temp <-  grid$longitude # stores longitude bounds of regions to 'temp'
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")

# Split region boundaries from 'cut()' into upper and lower longitudes
temp <- unlist(temp)
temp <- as.numeric(temp)
grid$max_long <- temp[2*1:nrow(grid)] # list of latitude upper bounds
grid$min_long <- temp[2*1:nrow(grid)-1] # list of latitude lower bounds
grid$avg_long <- 0.5*(grid$max_long + grid$min_long) # list of average region longitudes
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
timeplot <- ggplot(data = dat, aes(x = pickup_hour, y = count, col = region)) +
  geom_line(size = 1) +
  labs(title = "Analysis of 5 Most Popular Pickup and Dropoff Sites", 
       x = expression(italic("Hour of Day")), 
       y = expression(italic("Number of Rides"))) + 
  facet_wrap(~ type, ncol =1)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
MAXLAT1 <- 40.775
MINLAT1 <- 40.74
MAXLONG1 <- -73.965
MINLONG1<- -73.995
map <- get_map(location = c(MINLONG1 - 0.01 ,MINLAT1 - 0.01, MAXLONG1 + 0.01, 
                     MAXLAT1 + 0.01), source = "google", maptype = "roadmap") 

pop_map <- ggmap(map) +
  labs(x = expression(italic("Longitude")~"(degrees)"),
       y = expression(italic("Latitude")~"(degrees)")) +
  geom_label(aes(jitter(avg_long), jitter(avg_lat),label = regioncode), 
             data = grid, size = 2) + 
  theme(axis.line = element_blank(), 
    axis.text.x = element_blank(), 
    axis.text.y = element_blank(),
    axis.ticks = element_blank(), 
    axis.title.x = element_blank(), 
    axis.title.y = element_blank(),plot.margin=unit(c(0,0,0,0),"mm")) +
  scale_y_continuous(limits=c(MINLAT1, MAXLAT1)) + 
  scale_x_continuous(limits = c(MINLONG1, MAXLONG1))
```

```{r, warning = FALSE, message = FALSE, fig.width = 5, fig.height= 3}
grid.arrange(timeplot, pop_map, ncol = 2, widths = c(2,1))
```
Regions 1-5 are the most popular taxi pickup sites (1 = most popular, 5 = least). 
egions 6-10 are the most popular dropoff sites. Dropoff sites tend be most 
active at 7-8 AM. Pickup sites are most active at 8-9 PM. We will next 
investigate how popular pickup sites rank in terms of dropoff sites and vice 
versa.
  
* Which directions are taxis traveling on average at different times of the day? 
```{r, warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
# Remove rows with abnormal dropoff longitudes
rm(pos, grouped)
MAXLAT <- 40.85
MINLAT <- 40.675
MAXLONG <- -73.85
MINLONG <- -74.1
SIZEFACTOR  <-  0.015 # Desired Magnitude of Direction Arrows
GRID_RESOLUTION <- 25

pos$lat_region <- cut(pos$pickup_latitude,GRID_RESOLUTION) # cuts latitude into bins
pos$long_region <- cut(pos$pickup_longitude,GRID_RESOLUTION) # cuts latitude into bin
```

```{r, echo = FALSE, results = 'hide', warning = FALSE, message = FALSE}
# Add rows 'lat_dir' and 'long_dir' to the data frame 'pos' which respectively 
# contain the change in latitude and longitude for each ride.
pos <- mutate(pos, lat_dir = (dropoff_latitude - pickup_latitude), 
              long_dir = (dropoff_longitude - pickup_longitude)) 

# Create columns 'lat_dir_scaled' and 'long_dir_scaled', which contain the 
# coordinates of 'lat_dir' and 'long_dir' normalized so that the direction 
# vector has magnitude equal to 'SIZEFACTOR'
pos <- mutate(pos, lat_dir_scaled = lat_dir * SIZEFACTOR/sqrt(lat_dir^2 + long_dir^2), 
              long_dir_scaled = long_dir * SIZEFACTOR /sqrt(lat_dir^2 + long_dir^2)) 

# Note: NaN results for rides where pickup and dropoff locations are equal.

pos$daytime <- rep(NA, nrow(pos))

# Create new column of pos indicating whether pickup-time was in the morning or evening
pos$daytime <- ifelse(pos$pickup_hour < 12, "morning", "evening") 
pos$daytime <- factor(pos$daytime, levels= c("morning","evening"))
```

```{r,warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
# Groups data in 'pos' by 'lat_region' and 'long_region' (i.e. by region on the 
# grid). Calculates number of rides being picked up in each region ('count') and 
# the average direction traveled for each taxi rides beginning in each grid 
# (long_dir_avg, lat_dir_avg)
grouped <- group_by(pos, lat_region, long_region, daytime) %>% 
  summarize(lat_dir_avg = mean(lat_dir_scaled, na.rm = TRUE), 
            long_dir_avg = mean(long_dir_scaled, na.rm =TRUE), count = n()) 

temp <-  grouped$lat_region 
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")
temp <- unlist(temp)
temp <- as.numeric(temp)
grouped$max_lat <- temp[2*1:nrow(grouped)] # list of latitude upper bounds
grouped$min_lat <- temp[2*1:nrow(grouped)-1] # list of latitude lower bounds
grouped$avg_lat <- 0.5*(grouped$max_lat + grouped$min_lat) # list of average latitude
    
temp <-  grouped$long_region
temp <- gsub("\\(", "", temp)
temp <- gsub("\\]", "", temp)
temp <- strsplit(temp, ",")
temp <- unlist(temp)
temp <- as.numeric(temp)
grouped$max_long <- temp[2*1:nrow(grouped)] # list of latitude upper bounds
grouped$min_long <- temp[2*1:nrow(grouped)-1] # list of latitude lower bounds
grouped$avg_long <- 0.5*(grouped$max_long + grouped$min_long) # list of average longitude
```

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = 'hide'}
heatplot <- ggplot(data = grouped, aes(xmin = min_long, xmax = max_long, 
                   ymin = min_lat, ymax = max_lat, fill = count)) + 
  geom_rect() + 
  scale_fill_gradientn("Number of Pickups", 
                       colors = c("purple","blue","green","yellow", "orange", "red")) + 
  facet_wrap(~daytime)
```

```{r, warning = FALSE, message = FALSE, echo=FALSE, fig.width = 9, fig.height= 3}
MAXLAT1 <- 40.8
MINLAT1 <- 40.7
MAXLONG1 <- -73.92
MINLONG1<- -74.03
MINRIDES <-  30 # minimum number of rides for arrow to show up

map <- get_map(location = c(MINLONG, MINLAT, MAXLONG, MAXLAT), source = "stamen", 
               maptype = "watercolor", color = "bw")
```

```{r, warning = FALSE, message = FALSE, fig.width = 5.5, fig.height= 3}
ggmap(map) + geom_segment(aes(x = avg_long, xend = (avg_long + long_dir_avg), y = avg_lat, yend = (avg_lat + lat_dir_avg), color = count), data = filter(grouped, count > MINRIDES), arrow = arrow(length=unit(0.15, "cm"), ends="first", type = "closed"), size = 0.9) + labs(title = "Direction of Motion by Region", x = expression(italic("Longitude")~"(degrees)"), y = expression(italic("Latitude")~"(degrees)")) + scale_color_gradientn("Number of Rides", colors = c("purple","blue","green","yellow","orange","red")) + facet_wrap( ~ daytime) + scale_y_continuous(limits=c(MINLAT1, MAXLAT1)) + scale_x_continuous(limits = c(MINLONG1, MAXLONG1))
```
There is a general outward flux of taxis. Taxi rides starting farther from 
center tend to agree more on the outward direction (longer arrows). We will 
facet into more time catagories. We might investigate where outward directed 
taxis are going

* What is the average speed of taxis depending on the time of day and region of 
the city?
To do this, we would create a heat map of Manhattan where different colors 
indicate different travel speeds of taxis.

